Ionization of molecules by multicharged bared ions by using the stoichiometric model

### 2019-09-06 NewMonti24.tex
Table 2 modified

\documentclass[preprint,12pt]{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multirow}

@@

The damage caused by the impact of multicharged heavy projectile has 
become a field of interest because of its implementation in ion-beam 
cancer therapy. The effectiveness of the radiation depends on the 
choice of the ions. In particular, theoretical and experimental studies 
with different projectiles have concluded that charged carbon ions 
could be the most suitable ions to use. The study of such systems 
represent a challenge from the theoretical point of view. The most 
widely used method to compute ionization of multicharged atoms is the 
first Born approximation. 
This perturbative method warrants the $Z^{2}$ laws, where $Z$ is the 
projectile charge. However, the damage concentrated in the vicinities 
of the Bragg peak correspond to energies of hundreds of KeV/amu. 
Precisely in this region, often refered to as the intermediate energy 
region, the Born approximation starts to fail. Another theoretical 
issue arises from the targets themselves; we are dealing with complex 
molecules. This article deals precisely with these two aspects. 
First, we perform more appropriate calculations on the primary damage 
mechanism, i.e., atomic ionization by multicharged ions, which can 
replace the Born results. Second, we inspect and use an stoichiometric 
model, which reduces a molecule to a sum of atomic processes quantities 
weighted by the numbers of such atom in the molecule.

To overcome the first perturbative approximation limitations and 
since the projectiles are multicharged ions, we resort to the 
Continuum Distorted Wave-Eikonal Initial State (CDW), which includes 
higher perturbative corrections. We start from the premise that the 
ionization process is the mechanism that deposits the most significant 
amount of primary energy. This process produces an electron-energy 
spectrum that needs to be integrated. Moreover, the ejected electrons 
become a new source of local damage. These secondary electrons are 
included in Monte Carlo simulations and hence their behavior must be
investigated. With this end in mind, we calculate not only the 
ionization cross-sections but also energy and mean angular distributions 
of the emitted electrons.

@@

Then, by implementing the SSM and the CDW --instead of the first Born 
approximation--, we calculate ionization cross-section of several 
molecules (see Table~\ref{tab:families}) by the impact of antiprotons,
H$^{+}$, He$^{2+}$, Be$^{4+}$, C$^{6+}$,\ and O$^{8+}$. Furthermore, 
in Section 3, we investigate different DNA and RNA molecules
such as adenine, timine, cytosine, guanine, uracil, and also DNA 
backbone.

@@

It is well known that the residual electrons from the ionization 
process cause significant biological damage. To inspect this mechanism, 
in Section 3.3 and 3.4, we calculated the mean electron energy an 
angular distributions. Surprisingly, we found a substantial dependence 
of the charged projectile, which is unexpected in the first Born 
approximation.

@@

Lets us consider a molecule $M$ composed by $n_{\alpha}$ atoms of the
element $\alpha$, the SSM describes the total ionization cross section 
of the molecule $\sigma_{M}$ as a simple sum of ionization cross 
sections of the isolated atoms $\sigma_{\alpha}$, 
\begin{equation}
 \sigma_{M}=\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}\,.  
 \label{eq:sumion}
\end{equation}
We divided sixteen molecular targets of our interest in 3 families. 
The classification defined is given in Table~\ref{tab:families}.
\begin{table}
\begin{center}
\begin{tabular}{|p{0.06\textwidth}|p{0.55\textwidth}|}
\hline
 CH  & CH$_4$, C$_2$H$_2$, C$_2$H$_4$, C$_2$H$_6$, and C$_6$H$_6$ \\
\hline
 CHN & C$_5$H$_5$N, C$_4$H$_4$N$_2$, C$_2$H$_7$N, and CH$_5$N \\
\hline
 \multirow{4}{*}{DNA} & adenine (C$_5$H$_5$N$_5$), 
                        timine (C$_5$H$_6$N$_2$O$_2$), \\
     & cytosine (C$_4$H$_5$N$_2$O), guanine (C$_5$H$_5$N$_5$O), \\
     & uracil (C$_4$H$_4$N$_2$O$_2$), \\
     & DNA backbone (C$_5$H$_{10}$O$_5$P), \\
     & dry DNA (C$_{20}$H$_{27}$N$_7$O$_{13}$P$_2$)\\
\hline
\end{tabular}
\caption{Sixteen molecular targets of our interest classified in 3 
\label{tab:families}
families.}
\end{center}
\end{table}

In Figure 2, we report the total ionization cross sections by the 
impact of multicharged ions for the four fundamental DNA components: 
adenine, timine, cytosine, and guanine. Our reports for uracil and 
DNA backbone containing P are given in Figure 3. For adenine, the 
agreement with the experimental data available in our range of validity 
is quite good. For uracil, we have a puzzling situation: our results 
show good agreement with the experiments by proton impact of Itoh 
{\it et al.}~\cite{itoh2013}, but for the same target our theory fails 
by a factor of two for the impact of C$^{4+}$ and O$^{6+}$ ions. 
Nonetheless, it should be stated that our theoretical results coincide 
with the ones by Champion, Rivarola and 
collaborators~\cite{champion2012,agnihotri2012}, which may indicate a 
possible misstep of the experiments.

@@

As mentioned before, secondary electrons contribute to the total damage. 
Then, it is important not only the ejection energy but also the angle 
of emission. Once again, we can consider the single differential cross 
section in terms of the ejected electron solid angle $\Omega$, 
$d\sigma_{\alpha,nl}/d\Omega$, to be expressed as a distribution 
function, and define the mean angle $\overline{\theta}_{\alpha}$ as
\begin{equation}
\overline{\theta}_{\alpha}=\frac{\langle\theta_{\alpha}\rangle}
{\langle 1\rangle}=\frac{1}{\sigma_{\alpha}}\sum\limits_{nl}
\int d\Omega\,\theta\,\frac{d\sigma_{\alpha,nl}}{d\Omega}
\end{equation}
Figure 6 shows $\overline{\theta}_{\alpha}$ for our six elements of 
interest. A new important dependence of $\overline{\theta}_{\alpha}$ 
with $Z$ is observed. It is the common view~\cite{Rudd1992} that the 
angular dispersion of emitted electrons are nearly isotropic due to the
fact that angular anisotropy of sub-50-eV yield is not significant. 
A typical value of ejection angle used in the literature is 
$\overline{\theta}_{\alpha}\sim 70$ degrees~\cite{surdutovic2018}, and 
it is quite correct in the range of validity of the first Born 
approximation for any target. But when a distorted wave approximation 
is used, $\overline{\theta}_{\alpha}$ decreases substantially with $Z$ 
in the intermediate energy region, as observed in Figure 6. 
For example, for C$^{6+}$ impact, the Bragg peak occurs at 0.3 MeV/amu, 
where $\overline{\theta}_{\alpha}$ computed with the CDW method is 
about half of the value obtained with the first Born approximation. 
This correction should close the damage to the forward direction.

The only responsible of this correction is the capture to the continuum
effect: the larger the charge $Z$ the smaller $\overline{\theta}$ will 
be, of course at intermediate energies, not at high impact energies, 
where the Born approximation rules. One illustrative observation is the 
behavior of antiprotons: the projectile in this case repels the 
electrons making the distribution almost symmetric. Note the opposite 
effect of proton and antiprotons, they run one opposite to the other, 
as compared with the first Born approximation.

@@

One can correctly argue that\ the SSM considers an assembly of isolated
neutral atoms which is definitively unrealistic. A first improvement is to
consider that the atoms are not neutral and within the molecule they have
more or less electrons given by a \ charge $q_{\alpha }$. One parameter that
measures $q_{\alpha }$ is the Mulliken charge. This is not the only one,
there is a large variety of charges such as the net or natural atomic charge 
\cite{lee2003}, the lowin charge, etc.

Consider\ that the total amount of electrons $Q_{\alpha }\ $on element $%
\alpha $ are equally distributed on all the atoms $\alpha ,$ therefore each
element $\alpha \ $\ will have a an additional charge: $q_{\alpha
}=Q_{\alpha }/n_{\alpha }$. positive or negative depending on the relative
electronegative value respect to the other ones~\cite{rappe1991}. Instead
of the integer number of elements $n_{\alpha }$ of the atom $\alpha$, it
could be argued that we have now an continuous number of atoms given by 
\begin{equation}
n_{\alpha }^{\prime }=n_{\alpha }-\frac{q_{\alpha }}{v_{\alpha }^{CDW}}
\label{100}
\end{equation}%
In the case of neutral atoms $q_{\alpha }=0,$ we obviously recover $%
n_{\alpha }^{\prime }=n_{\alpha }$ as before$.$

To inspect the effect of the $q_{\alpha }$ we have run GAMES with the 
6-31G basis set, including polarization functions for all the atoms. 
The calculations were carried out implementing the B3LYP 
functional~\cite{Becke1993,Stephens1994}. In Table~(\ref{110}) we display
$q_{\alpha }$ of the different DNA basis and in the last column we rewrite
the new stoichiometric expression. 
\begin{equation}
\begin{array}{|l|l|l|l|l|l|}
\hline
\text{Element}\,\,\, & \mathrm{H} & \mathrm{C} & \mathrm{N} & \mathrm{O} & 
\text{New stoichiometry} \\ 
\hline
\mathrm{Adenine}  & +0.32 & -0.55 & +0.23 &       & 
\mathrm{C}_{4.92}\mathrm{H}_{5.14}\mathrm{N}_{4.77} \\ 
\hline
\mathrm{Thymine}  & +0.20 & -0.54 & +0.19 & -0.52 & 
\mathrm{C}_{4.95}\mathrm{H}_{6.13}\mathrm{N}_{1.95}\mathrm{O}_{2.13} \\ 
\hline
\mathrm{Cytosine} & +0.28 & -0.56 & +0.21 & -0.53 & 
\mathrm{C}_{3.93}\mathrm{H}_{5.14}\mathrm{N}_{2.79}\mathrm{O}_{1.13} \\ 
\hline
\mathrm{Guanine}  & +0.46 & -0.58 & +0.20 & -0.36 & 
\mathrm{C}_{4.89}\mathrm{H}_{5.15}\mathrm{N}_{4.80}\mathrm{O}_{1.09} \\ 
\hline
\end{array}
\label{110}
\end{equation}

All the previous formula holds with the simple replacement of integer value $%
n_{\alpha }$ by the continuous values\ $n_{\alpha }^{\prime }$ to produce a
new ionization cross section $\sigma ^{\prime }$

In Figure 7, we displays the relative difference of ionization cross
sections 
\begin{equation}
e=\frac{\sigma ^{\prime }-\sigma }{\sigma }\times 100  \label{120}
\end{equation}%
for the four DNA basis. The difference is \ just very few percents.
Indicating that the SSM is a quite robust model to treat the type of
molecule we are dealing with, and of course within the range error that one
can expect from the SSM.

>>>>

\documentclass[preprint,12pt]{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{gensymb}
\usepackage{float}

@@

The damage caused by the impact of multicharged heavy projectiles has 
become a field of interest due to its recent implementation in ion-beam 
cancer therapy. The effectiveness of the radiation depends on the 
choice of the ions. In particular, theoretical and experimental studies 
with different projectiles have concluded that charged carbon ions 
could be the most suitable ions to use. The study of such systems 
represent a challenge from the theoretical point of view. The most 
widely used method to compute ionization of multicharged atoms is the 
first Born approximation. 
This perturbative method warrants the $Z^{2}$ laws, where $Z$ is the 
projectile charge. However, the damage concentrated in the vicinities 
of the Bragg peak correspond to energies of hundreds of KeV/amu. 
Precisely in this region, often refered to as the intermediate energy 
region, the Born approximation starts to fail. Another theoretical 
issue arises from the targets themselves; we are dealing with complex 
molecules. This article deals precisely with these two aspects. 
First, we perform more appropriate calculations on the primary damage 
mechanism, i.e., atomic ionization by multicharged ions, which can 
replace the Born results. Second, we inspect and use an stoichiometric 
model, which reduces a molecule to a sum of atomic processes quantities 
weighted by the number of such atom in the molecule.

To overcome the first perturbative approximation limitations, and 
since the projectiles are multicharged ions, we resort to the 
Continuum Distorted Wave-Eikonal Initial State (CDW), which includes 
higher perturbative corrections. We start from the premise that the 
ionization process is the mechanism that deposits the most significant 
amount of primary energy. This process produces an electron-energy 
spectrum that needs to be integrated. Moreover, the ejected electrons 
become a new source of local damage. These secondary electrons are 
included in Monte Carlo simulations and hence their behavior must be
investigated. With this end in mind, we calculate not only the 
ionization cross-sections but also energy and mean angular distributions 
of the emitted electrons.

@@

Then, by implementing the SSM and the CDW --instead of the first Born 
approximation--, we calculate ionization cross-section of several 
molecules (see Table~\ref{tab:families}) by the impact of antiprotons,
H$^{+}$, He$^{2+}$, Be$^{4+}$, C$^{6+}$,\ and O$^{8+}$. Furthermore, 
in Section 3, we investigate different DNA and RNA molecules
such as adenine, cytosine, guanine, thymine, uracil, and also DNA 
backbone.

@@

It is well known that the residual electrons from the ionization 
process cause significant biological damage. To inspect this mechanism, 
in Section 3.3 and 3.4, we calculated the mean electron energy and
angular distributions. Surprisingly, we found a substantial dependence 
of the charged projectile, which is unexpected in the first Born 
approximation.

@@

Lets us consider a molecule $M$ composed by $n_{\alpha}$ atoms of the
element $\alpha$, the SSM describes the total ionization cross section 
of the molecule $\sigma_{M}$ as a simple sum of ionization cross 
sections of the isolated atoms $\sigma_{\alpha}$, 
\begin{equation}
 \sigma_{M}=\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}\,.  
 \label{eq:sumion}
\end{equation}
We divided sixteen molecular targets of our interest in three families. 
The classification defined is given in Table~\ref{tab:families}.
\begin{table}
\begin{center}
\begin{tabular}{|p{0.06\textwidth}|p{0.65\textwidth}|}
\hline
 CH  & CH$_4$, C$_2$H$_2$, C$_2$H$_4$, C$_2$H$_6$, and C$_6$H$_6$ \\
\hline
 CHN & C$_5$H$_5$N, C$_4$H$_4$N$_2$, C$_2$H$_7$N, and CH$_5$N \\
\hline
 \multirow{4}{*}{DNA} & adenine (C$_5$H$_5$N$_5$), 
                        cytosine (C$_4$H$_5$N$_2$O), \\
     & guanine (C$_5$H$_5$N$_5$O), thymine (C$_5$H$_6$N$_2$O$_2$), \\
     & uracil (C$_4$H$_4$N$_2$O$_2$), DNA backbone (C$_5$H$_{10}$O$_5$P), \\
     & dry DNA (C$_{20}$H$_{27}$N$_7$O$_{13}$P$_2$)\\
\hline
\end{tabular}
\caption{Sixteen molecular targets of our interest classified in three 
families.}
\label{tab:families}
\end{center}
\end{table}

In Figure 2, we report the total ionization cross sections by the 
impact of multicharged ions for the four fundamental DNA components: 
adenine, cytosine, guanine and thymine. Our reports for uracil and 
DNA backbone containing P are given in Figure 3. For adenine, the 
agreement with the experimental data available in our range of validity 
is quite good. For uracil, we have a puzzling situation: our results 
show good agreement with the experiments by proton impact of Itoh 
{\it et al.}~\cite{itoh2013}, but for the same target our theory fails 
by a factor of two for the impact of C$^{4+}$ and O$^{6+}$ ions. 
Nonetheless, it should be stated that our theoretical results coincide 
with calculations by Champion, Rivarola and 
collaborators~\cite{champion2012,agnihotri2012}, which may indicate a 
possible misstep of the experiments.

@@

As mentioned before, secondary electrons contribute to the total damage. 
Then, not only is the ejection energy important but also the angle 
of emission. Once again, we can consider the single differential cross 
section in terms of the ejected electron solid angle $\Omega$, 
$d\sigma_{\alpha,nl}/d\Omega$, to be expressed as a distribution function, 
and the mean angle $\overline{\theta}_{\alpha}$ can be defined as
\begin{equation}
\overline{\theta}_{\alpha}=\frac{\langle\theta_{\alpha}\rangle}
{\langle 1\rangle}=\frac{1}{\sigma_{\alpha}}\sum\limits_{nl}
\int d\Omega\,\theta\,\frac{d\sigma_{\alpha,nl}}{d\Omega}
\end{equation}
Figure 6 shows $\overline{\theta}_{\alpha}$ for our six elements of 
interest. A new important dependence of $\overline{\theta}_{\alpha}$ 
with $Z$ is observed. It is a general belief~\cite{Rudd1992} that the 
angular dispersion of emitted electrons are nearly isotropic. This effect 
is caused by the insignificant angular anisotropy of sub-50-eV yield. 
A typical value for the ejection angle considered in the literature is 
$\overline{\theta}_{\alpha}\sim 70\degree$~\cite{surdutovic2018}, and 
it is quite correct in the range of validity of the first Born 
approximation for any target. But, when a distorted wave approximation 
is used, $\overline{\theta}_{\alpha}$ decreases substantially with $Z$ 
in the intermediate energy region, as observed in Figure 6. 
For example, for C$^{6+}$ impact, the Bragg peak occurs at 0.3 MeV/amu, 
where $\overline{\theta}_{\alpha}$, computed with the CDW method, is 
about half of the value obtained with the first Born approximation. 
This correction should close the damage to the forward direction.

We can attribute this correction to the capture to the continuum effect;
the larger the charge $Z$, the smaller $\overline{\theta}$ will be. Of 
course, this effect only holds at intermediate energies and not at high 
impact energies, where the Born approximation rules. One illustrative 
observation is the behavior of antiprotons: the projectile in this case 
repels the electrons, making the distribution almost symmetric. 
Note the opposite effect of proton and antiprotons; they run one 
opposite to the other, as compared with the first Born approximation.

@@

The SSM considers the molecule to be assembled by isolated neutral atoms, 
which is definitively unrealistic. A first improvement can be suggested 
by assuming that the atoms are not neutral and that they have an uneven
distribution of electrons within the molecule; this distribution can be
given by an effective charge $q_{\alpha}$. A possible value for 
$q_{\alpha}$ is given by the Mulliken charge. However, there are a wide
variety of charge distributions, such as the net or natural atomic
charge~\cite{lee2003}, the L\"owdin charge, etc.

Consider that the total amount of electrons $Q_{\alpha }$ on the element
$\alpha$ are equally distributed on all the $\alpha$ atoms. Therefore, 
each element $\alpha$ will have an additional charge: 
$q_{\alpha}=Q_{\alpha}/n_{\alpha}$, which can be positive or negative.
This value will depend on the relative electronegativity value with 
respect to the other atoms~\cite{rappe1991}. Now, instead of an
integer number of elements $n_{\alpha}$ of the atom $\alpha$, we have a 
fractional number of atoms given by 
\begin{equation}
n_{\alpha }^{\prime }=n_{\alpha }-
\frac{q_{\alpha }}{\nu_{\alpha }^{\text{CDW}}}
\label{eq:newstoi}
\end{equation}%
In the case of neutral atoms, $q_{\alpha}=0$ and we recover 
$n_{\alpha}^{\prime}=n_{\alpha}$, as it should be.

To inspect the effect of $q_{\alpha}$, we computed the molecular
structure of several nucleobases with the {\sc gamess} code. We 
used the 6-31G(d,p) basis set, which includes polarization 
functions for all the atoms. The calculations were carried out 
implementing the B3LYP functional~\cite{Becke1993,Stephens1994} to 
account for the correlation and exchange effects. 
In Table~\ref{tab:newstoi}, we display the charge $q_{\alpha}$ of four 
DNA molecules. The last column shows the new stoichiometric formula
found by implementing Eq.~(\ref{eq:newstoi}).
\begin{table}[H]
\begin{center}
\begin{tabular}{|p{0.12\textwidth}|p{0.07\textwidth}|p{0.07\textwidth}|p{
0.07\textwidth}|p{0.07\textwidth}|p{0.25\textwidth}|}
\hline
Element & C & H & N & O & New stoichiometry \\
\hline
Adenine & +0.32 & +0.23 & --0.55 &       & C$_{4.92}$H$_{4.77}$N$_{5.14}$ \\ 
\hline
Cytosine & +0.28 & +0.21 & --0.56 & --0.53 & C$_{3.93}$H$_{2.79}$N$_{5.14}$O$_{1.13}$ \\ 
\hline
Guanine & +0.46 & +0.20 & --0.58 & --0.36 & C$_{4.89}$H$_{4.80}$N$_{5.15}$O$_{1.09}$ \\ 
\hline
Thymine & +0.20 & +0.19 & --0.54 & --0.52 & C$_{4.95}$H$_{1.95}$N$_{6.13}$O$_{2.13}$ \\ 
\hline
\end{tabular}
\caption{Effective charge $q_{\alpha}$ and new stoichiometric formula 
defined by Eq.~(\ref{eq:newstoi}) for four DNA molecules.}
\label{tab:newstoi}
\end{center}
\end{table}

The previous formula can be rewritten by replacing the integer value 
$n_{\alpha}$ by the fractional values $n_{\alpha}^{\prime}$. This 
analog formulations are now implemented to compute the new ionization 
cross sections, $\sigma^{\prime}$.

In Figure 7, we display the relative difference of ionization cross 
sections 
\begin{equation}
e=\frac{\sigma^{\prime}-\sigma}{\sigma}\times 100
\label{120}
\end{equation}
for the four DNA basis. The difference obtained is of just very few 
percents, which indicates that the SSM is a quite robust model to treat 
the type of molecule we are dealing with, and of course within the 
range error that one can expect from the SSM.
